---
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
```

## mult_sampsize

## [![contributions welcome](https://img.shields.io/badge/contributions-welcome-brightgreen.svg?style=flat)](https://github.com/dwyl/esta/issues)

## Description 

R package for determining sample size required in studies using multivariate mixed continuous and discrete endpoints, when analysed using a latent variable model. The package facilitates sample size estimation for mixed outcomes that are combined to form either composite, co-primary and multiple primary endpoints.

## Getting started 

Install from Github using `devtools::install_github("martinamcm/mult_sampsize")`

Implementation as a Shiny app with further documentation on functionality and examples available at [MultSampSize](https://github.com/martinamcm/MultSampSize)

## Details 

### Composite Endpoints 

Function `sampsizecomp()` provides the required sample size using both the augmented binary method
based on a latent variable model and a standard binary method based on a logistic regression model
with the following inputs:

 * `augmean` mean risk difference treatment effect estimated using augmented binary method
 * `binmean` mean risk difference treatment effect estimated using standard binary method
 * `augvar` variance of risk difference treatment effect estimated using augmented binary method
 * `binvar` variance of risk difference treatment effect estimated using standard binary method
 * `alpha` one-sided alpha level 
 * `beta` beta level: 1-desired power 

Estimates of these quantities can be obtained from existing data using the `augbin_rheum` package, as shown below.

#### Example

Assuming the endpoint of interest is a composite endpoint comprised of two continuous and one binary component, 
the sample size required in each arm when the dichotomisation thresholds are equal to 18 and 6 is obtained as below. 

More details and `Egdata21` can be obtained from [MultSampSize](https://github.com/martinamcm/MultSampSize). 

```{r, eval=F}

devtools::install_github("martinamcm/augbin_rheum")

data_fit <- augbinrheum(Egdata21,2,1,c(18,6))

augmean_est <- data_fit$risk_diff$est[1]
binmean_est <- data_fit$risk_diff$est[2]

augvar_est <- ((data_fit$risk_diff$ci_upper[1]-augmean_est)/1.96)^2
binvar_est <- ((data_fit$risk_diff$ci_upper[2]-binmean_est)/1.96)^2

sampsizecomp(augmean_est,binmean_est,augvar_est,binvar_est,0.05,0.2)

```


### References

McMenamin M, Barrett JK, Berglind A, Wason JMS. Sample Size Estimation using a Latent Variable Model for Mixed Outcome Co-Primary, Multiple Primary and Composite Endpoints. *arXiv*. 2019. [arXiv:1912.05258](https://arxiv.org/abs/1912.05258).

McMenamin M, Grayling MJ, Berglind A, Wason JMS. Increasing power in the analysis of responder endpoints in rheumatology: a software tutorial. *medRxiv*. 2020. doi: [10.1101/2020.07.28.20163378](https://www.medrxiv.org/content/10.1101/2020.07.28.20163378v1)

McMenamin M, Barrett JK, Berglind A, Wason JM. Employing a latent variable framework to improve efficiency in composite endpoint analysis. *Statistical Methods in Medical Research*. 2021;30(3):702-716. doi: [10.1177/0962280220970986](https://doi.org/10.1177/0962280220970986)